paper2_resistance_mechanisms_mixtures

### 2016-06-28 inst/documents/paper2_resistance_mechanisms_mixtures.Rmd
putting figs side by side

#### (This file is automagically generated by R from paper2_resistance_mechanisms_mixtures.Rmd)
#### Target : Malaria journal

@@

This paper explores the mechanisms by which insectide resistance is expected to increase with the use of insecticides in isolation, sequence/rotation and mixtures. It is aimed at a readership with limited experience of population genetics and resistance. It uses a model representing two independent loci coding for resistance to two insecticides. It explores factors that are expected to influence the development of resitance, notably the starting frequency of resistance alleles, their dominance, the exposure of insects to the insecticides, the effectiveness of the insecticides, the selective adavantage of resistance and the dominance of the alleles. There is uncertainty about the values of these parameters in field situations. We demonstrate how the values of these parameters are expected to influence the development of resistance and the relative benefits of alternative insecticide application strategies. We describe an R package and user-interface that allows further exploration of the development of resistance. The user-interface (https://andysouth.shinyapps.io/MixSeqResist1) allows replication of our results and exploration of what happens when inputs are changed outside of the ranges we have used.

@@

  print( ggplot(df_resist, aes(x=generation, y=resistance, colour=exposure)) + 
         labs(colour = "exposure") +
         coord_trans(y = "log10") +    
         geom_point() #show.legend=FALSE) +

@@

  print( ggplot(df_resist, aes(x=generation, y=resistance, colour=h.RS1_A0)) + 
         labs(colour = "dominance") +
         coord_trans(y = "log10") + 
         geom_point() #show.legend=FALSE) +

@@

  print( ggplot(df_resist, aes(x=generation, y=resistance, colour=rr_restoration_ins1)) + 
         labs(colour = "selective advantage") +
         coord_trans(y = "log10") +   
         geom_point() #show.legend=FALSE) +

@@

With all control inputs set equal at 0.5 resistance arises slower for sequential than mixture (Fig. 1). Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time. When used in sequence the curve for each insecticide individually is steeper but, because they happen one after the other, it takes longer to reach the resistance threshold for both.

```{r, eval=TRUE, echo=FALSE, fig.height=4}
runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.5 , h.RS2_0B = 0.5 , exposure = 0.5 , phi.SS1_A0 = 0.5 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2') )
cat("Fig. 1. All control inputs equal at 0.5")
```

@@

In a mixture the mechanism is that a more effective insecticide provides greater protection to the other. For example taking the previous scenario and increasing the effectiveness of one insecticide from 0.5 to 0.9 increases the time-to-resistance for the 2nd insecticide from around 45 to 75 generations (Fig. 2). In the plot you can see that resistance to insecticide1 (with the increased effectiveness) rises at a similar speed to Fig.2. In contrast resistance to insecticide2 increases more slowly initially. Once insecticide1 reaches it's resistance threshold (around 40 generations), the curve for insecticide2 becomes steeper. This can be explained by the mechanism that once insecticide1 has become ineffective due to resistance it stops killing indivduals that are resistant to insecticide2 and thus stops 'protecting' it from it's own rise in resistance.

```{r, eval=TRUE, echo=FALSE, fig.height=4}
runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.5 , h.RS2_0B = 0.5 , exposure = 0.5 , phi.SS1_A0 = 0.9 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2') )
cat("Fig. 2. Effectiveness of insecticide1 increased from 0.5 to 0.9")
```

When the effectiveness of both insecticides are increased to 0.9 the time-to-resistance is increased still further to around 125 generations (Fig. 3). In this case the curves for the rise in resistance to each insecticide are identical. The curves are shallower than previous plots because individuals that are resistant to each insecticide are being killed by the other thus reducing the increase in resistance alleles. As the resistance frequencies go above 10% around generation 100 both insecticides provide less protection to the other and the slope of the resistance curve increases.

```{r, eval=TRUE, echo=FALSE, fig.height=4}
runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.5 , h.RS2_0B = 0.5 , exposure = 0.5 , phi.SS1_A0 = 0.9 , phi.SS2_0B = 0.9 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2') )
cat("Fig. 3. Effectiveness of both insecticides 0.9")
```

**Dominance of resistance alleles**
Increasing the dominance of the alleles within a mixture has the same direction of effect as it does for sole use. Increasing the dominance of the allele corresponding to insecticide1 from 0.5 to 0.9 decreases the time-to-resistance to insecticide1 from 40 to 30 generations but leaves the time-to-resistance for the mixture the same (Fig. 4). Setting the dominance for bothe alleles to 0.9 reduces the time-to-resistance for the mixture to 30 generations (Fig. 5). In both of these cases sequential use continues to have a longer time-to-resistance.

```{r, eval=TRUE, echo=FALSE, fig.height=4}
runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.9 , h.RS2_0B = 0.5 , exposure = 0.5 , phi.SS1_A0 = 0.5 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2') )
cat("Fig. 4. Dominance of insecticide1 0.9")
```

```{r, eval=TRUE, echo=FALSE, fig.height=4}
runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.9 , h.RS2_0B = 0.9 , exposure = 0.5 , phi.SS1_A0 = 0.5 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2') )
cat("Fig. 5. Dominance of both insecticides 0.9")
```

>>>>

(This file is automagically generated by R from paper2_resistance_mechanisms_mixtures.Rmd)
Target : Malaria journal

#### TO DO
1. decide whether to use selective advantage or rr restoration coefficient

@@

This paper explores the mechanisms by which insectide resistance is expected to increase with the use of insecticides in isolation, sequence/rotation and mixtures. It is aimed at a readership with limited experience of population genetics and resistance. It uses a model representing two independent loci coding for resistance to two insecticides. It explores factors that are expected to influence the development of resitance, notably the starting frequency of resistance alleles, their dominance, the exposure of insects to the insecticides, the effectiveness of the insecticides, the selective advantage of resistance and the dominance of the alleles. There is uncertainty about the values of these parameters in field situations. We demonstrate how the values of these parameters are expected to influence the development of resistance and the relative benefits of alternative insecticide application strategies. We describe an R package and user-interface that allows further exploration of the development of resistance. The user-interface (https://andysouth.shinyapps.io/MixSeqResist1) allows replication of our results and exploration of what happens when inputs are changed outside of the ranges we have used.

@@

  print( ggplot(df_resist, aes(x=generation, y=resistance, colour=exposure)) + 
         theme_bw() +
         labs(colour = "exposure") +
         coord_trans(y = "log10") +    
         geom_point() #show.legend=FALSE) +

@@

  print( ggplot(df_resist, aes(x=generation, y=resistance, colour=h.RS1_A0)) + 
         theme_bw() +
         labs(colour = "dominance") +
         coord_trans(y = "log10") + 
         geom_point() #show.legend=FALSE) +

@@

  print( ggplot(df_resist, aes(x=generation, y=resistance, colour=rr_restoration_ins1)) + 
         theme_bw() +
         labs(colour = "selective advantage") +
         coord_trans(y = "log10") +   
         geom_point() #show.legend=FALSE) +

@@

With all control inputs set equal at 0.5 resistance arises slower for sequential than mixture (Fig. 1 A). Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time. When used in sequence the curve for each insecticide individually is steeper but, because they happen one after the other, it takes longer to reach the resistance threshold for both.

@@

```{r, eval=TRUE, echo=FALSE, fig.height=3.7, fig.width=3.7, fig.show='hold', fig.align='left', fig.cap="A. All control inputs equal at 0.5, B. Effectiveness of insecticide1 increased from 0.5 to 0.9"}
#fig.cap seems necessary to keep figs side by side

#plot margins
oldpar <- par(mar=c(4, 4, 2, 0)) #b,l,t,r default c(5, 4, 4, 2)

runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.5 , h.RS2_0B = 0.5 , exposure = 0.5 , phi.SS1_A0 = 0.5 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2'), addLegend=FALSE, main='A.' )

#plot margins
par(mar=c(4, 0, 2, 0)) #b,l,t,r default c(5, 4, 4, 2)

runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.5 , h.RS2_0B = 0.5 , exposure = 0.5 , phi.SS1_A0 = 0.9 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2'), ylab="", main='B.' )

par(oldpar)

In a mixture the mechanism is that a more effective insecticide provides greater protection to the other. For example taking the previous scenario and increasing the effectiveness of one insecticide from 0.5 to 0.9 increases the time-to-resistance for the 2nd insecticide from around 45 to 75 generations (Fig. 1 B). In the plot you can see that resistance to insecticide1 (with the increased effectiveness) rises at a similar speed to Fig. 1A. In contrast resistance to insecticide2 increases more slowly initially. Once insecticide1 reaches it's resistance threshold (around 40 generations), the curve for insecticide2 becomes steeper. This can be explained by the mechanism that once insecticide1 has become ineffective due to resistance it stops killing indivduals that are resistant to insecticide2 and thus stops 'protecting' it from it's own rise in resistance.

When the effectiveness of both insecticides are increased to 0.9 the time-to-resistance is increased still further to around 125 generations (Fig. 2). In this case the curves for the rise in resistance to each insecticide are identical. The curves are shallower than previous plots because individuals that are resistant to each insecticide are being killed by the other thus reducing the increase in resistance alleles. As the resistance frequencies go above 10% around generation 100 both insecticides provide less protection to the other and the slope of the resistance curve increases.

```{r, eval=TRUE, echo=FALSE, fig.height=3.7, fig.width=3.7, fig.show='hold', fig.align='left', fig.cap="Effectiveness of both insecticides 0.9"}

#plot margins
oldpar <- par(mar=c(4, 4, 2, 0)) #b,l,t,r default c(5, 4, 4, 2)

runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.5 , h.RS2_0B = 0.5 , exposure = 0.5 , phi.SS1_A0 = 0.9 , phi.SS2_0B = 0.9 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2') )

**Dominance of resistance alleles**
Increasing the dominance of the alleles within a mixture has the same direction of effect as it does for sole use. Increasing the dominance of the allele corresponding to insecticide1 from 0.5 to 0.9 decreases the time-to-resistance to insecticide1 from 40 to 30 generations but leaves the time-to-resistance for the mixture the same (Fig. 3A). Setting the dominance for both alleles to 0.9 reduces the time-to-resistance for the mixture to 30 generations (Fig. 3B). In both of these cases sequential use continues to have a longer time-to-resistance.

```{r, eval=TRUE, echo=FALSE, fig.height=3.7, fig.width=3.7, fig.show='hold', fig.align='left', fig.cap="A. Dominance of insecticide1 0.9, B. Dominance of both insecticides 0.9"}

#plot margins
oldpar <- par(mar=c(4, 4, 2, 0)) #b,l,t,r default c(5, 4, 4, 2)

runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.9 , h.RS2_0B = 0.5 , exposure = 0.5 , phi.SS1_A0 = 0.5 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2'), addLegend=FALSE, main='A.' )
#cat("Dominance of insecticide1 0.9")

#plot margins
par(mar=c(4, 0, 2, 0)) #b,l,t,r default c(5, 4, 4, 2)

runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.9 , h.RS2_0B = 0.9 , exposure = 0.5 , phi.SS1_A0 = 0.5 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('seq','','adapt','mix2'), ylab='', main='B.' )
#cat("Dominance of both insecticides 0.9")

par(oldpar)
